---
title: "Climate Summary for Canyonlands National Park"
author: "Matthew Van Scoyoc"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r Setup, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
#-- Packages
# install.packages(c("tidyverse", "lubridate", "cowplot", "knitr"))
library("tidyverse") # Data manipulation
library("lubridate") # Easily work with date/time data
library("cowplot") # Streamlining for ggplot2
library("knitr") # Makes nice tables in output

#-- Custom functions --
#-- Create temp dataframes
temps.dat.fn <- function(wx.dat, my.dist, last.day){
  
  #-- Subset temp data
  temp.dat <- wx.dat %>%
    filter(element != "prcp" & 
           district == my.dist & date <= last.day) %>%
    mutate(degC = value / 10) %>%  # Convert to Celsius
    select(-value)
  #-- 30-yr Temp Stats (1981-2010 water years)
  temps.30yr <- temp.dat %>%
    filter(waterYr >= 1981 & waterYr <= 2010) %>%
    group_by(element, district, waterMonth) %>%
    summarise(mean.30yr = mean(degC, na.rm = T)) %>%
    left_join(month.df) %>%
    select(element, district, mthName, mean.30yr)
  #-- Monthly Temp Departures
  temps.depart <- temp.dat %>%
    select(parkUnit, district, waterYr, date, month, element, degC) %>%
    left_join(month.df) %>%
    group_by(element, district, waterYr, mthName) %>%
    summarise(mean.degC = mean(degC, na.rm = T)) %>%
    left_join(temps.30yr) %>%
    mutate(departure = mean.degC - mean.30yr,
           year = ifelse(mthName %in% c("Oct", "Nov", "Dec"), waterYr - 1, waterYr), 
           mid.date = ymd(paste(year, mthName, 15, sep = "-"))) %>%
    arrange(element, mid.date)
  #-- Rank departures
  temps.rank <- temps.depart %>%
    group_by(element, district, mthName) %>%
    mutate(warmest = min_rank(-mean.degC),
           coolest = min_rank(mean.degC)) %>%
    arrange(element, district, mthName, warmest)
  x <- list(temp.dat, temps.30yr, temps.depart, temps.rank)
  names(x) <- c("dat", "Avg30yr", "depart", "rank")
  return(x)
}

#-- Create temp figure
temp.fig <- function(dat){
  t1 <- ggplot(dat, aes(x = mid.date, y = departure)) +
    geom_bar(data = filter(dat, departure >= 0), stat = "identity",
             position = position_dodge(), color = "red", fill = "red") +
    geom_bar(data = filter(dat, departure < 0), stat = "identity",
             position = position_dodge(), color = "blue", fill = "blue") +
    geom_hline(aes(yintercept = 0), linetype="solid", size = 0.5, color = "black") +
    facet_wrap(~ element, nrow = 2, ncol = 1, scales = "free_y",
               labeller = as_labeller(c(tmax = "Daily maximum temperature (TMAX)", 
                                        tmin = "Daily minimum temperature (TMIN)"))) +
    labs(x = "Year", y = "Departure (°C)", title = "Departure of avg. monthly temps") +
    my.theme
  t2 <- ggplot(dat, aes(x = mthName, y = departure, group = waterYr)) +
    geom_line(stat = "identity", color = "gray") +
    geom_text(data = parkUnit, mapping = aes(x = Inf, y = -Inf, label = label), 
              color = "gray40", hjust = 1.2, vjust = -1) +
    geom_line(data = filter(dat, waterYr == water.yr), stat = "identity", 
              color = "red3", size = 1.5) +
    geom_text(data = parkUnit, mapping = aes(x = Inf, y = Inf, label = waterYr), 
              color = "red3", hjust = 2, vjust = 1.5) +
    geom_hline(aes(yintercept = 0), linetype="dashed", size = 0.75, color = "black") +
    facet_wrap(~ element, nrow = 2, ncol = 1,  scales = "free_y",
               labeller = as_labeller(c(tmax = "Daily maximum temperature (TMAX)", 
                                        tmin = "Daily minimum temperature (TMIN)"))) +
    labs(title = paste("WY", water.yr, "relative to other years", sep = " ")) +
    my.theme +
    theme(axis.title.y = element_blank(), 
          axis.text.x  = element_text(angle=45, vjust=0.5))
  return(multiplot(t1, t2, cols = 2))
}

#-- Create PRCP dataframes
prcp.dat.fn <- function(wx.dat, my.dist, last.day){
  #-- Subset PRCP data
  prcp.dat <- wx.dat %>%
    filter(element == "prcp" & 
             district == my.dist & date <= last.day) %>%
    mutate(PRCPmm = value / 10, 
           PRCPin = PRCPmm * (1/25.4)) %>% # Convert to inches
    select(-value)
  #-- 30-yr PRCP Stats (1981-2010 water years)
  prcp.30yr = prcp.dat %>%
    filter(waterYr >= 1981 & waterYr <= 2010) %>%
    group_by(district, waterYr, mthName) %>%
    summarise(element = "mth", 
              Obs = sum(PRCPmm, na.rm = T))
  wy.30yr <- prcp.30yr %>%
    mutate(element = "wy", 
           Obs = cumsum(Obs))
  prcp.30yr <- bind_rows(prcp.30yr, wy.30yr) %>%
    group_by(district, mthName, element) %>%
    summarise(mean.30yr = mean(Obs), 
              sd.30yr = sd(Obs))
  #- Calculate percent average and departure
  # Calculate monthly precip totals
  mth.prcp <- prcp.dat %>%
    select(district, year, waterYr, mthName, date, PRCPmm) %>%
    group_by(district, year, waterYr, mthName) %>%
    summarise(element = "mth", 
              Obs = sum(PRCPmm, na.rm = T))
  # Calculate water year precip totals
  wy.prcp <- mth.prcp %>%
    mutate(element = "wy", 
           Obs = cumsum(Obs))
  # Combine dataframes and calculate pct avg and departure
  pct.prcp <- bind_rows(mth.prcp, wy.prcp) %>%
    left_join(prcp.30yr) %>%
    mutate(pct.avg = (Obs / mean.30yr) * 100,
           departure = Obs - mean.30yr,
           mid.date = ymd(paste(year, mthName, 15, sep = "-")))
  # Rank monthly accumulations by water year.
  wy.rank <- wy.prcp %>%
    group_by(district, mthName) %>%
    mutate(Wettest = min_rank(-Obs),
           Driest = min_rank(Obs))
  x <- list(prcp.dat, prcp.30yr, pct.prcp, wy.rank)
  names(x) <- c("dat", "Avg30yr", "pct", "rank")
  return(x)
}

#-- Create PRCP figure
prcp.fig <- function(dat){
  p1 <- ggplot(dat, aes(x = mid.date, y = pct.avg)) +
    geom_bar(data = filter(dat, pct.avg >= 100), stat = "identity",
             position = position_dodge(), color = "forestgreen", fill = "forestgreen") +
    geom_bar(data = filter(dat, pct.avg < 100), stat = "identity",
             position = position_dodge(), color = "chocolate3", fill = "chocolate3") +
    geom_hline(aes(yintercept = 100), linetype="dashed", size = 0.75, 
               color = "black") +
    facet_wrap(~ element, nrow = 2, ncol = 1,  scales = "free_y",
               labeller = as_labeller(c(mth = "Monthly precipitation totals", 
                                        wy = "Water year accumulation"))) +
    labs(x = "Year", y = "Percent (%)", title = "Percent avg. monthly precip.") +
    my.theme
  p2 <- ggplot(dat, aes(x = mthName, y = pct.avg, group = waterYr)) +
    geom_line(stat = "identity", color = "gray") +
    geom_text(data = parkUnit, mapping = aes(x = -Inf, y = Inf, label = label), 
              color = "gray40", hjust = -0.5, vjust = 1.5) +
    geom_line(data = filter(dat, waterYr == water.yr), stat = "identity", 
              color = "forestgreen", size = 1.5) +
    geom_text(data = parkUnit, mapping = aes(x = Inf, y = Inf, label = waterYr), 
              color = "forestgreen", hjust = 1.25, vjust = 1.5) +
    geom_hline(aes(yintercept = 100), linetype="dashed", size = 0.75, color = "black") +
    facet_wrap(~ element, nrow = 2, ncol = 1,  scales = "free_y",
               labeller = as_labeller(c(mth = "Monthly precipitation totals", 
                                        wy = "Water year accumulation"))) +
    labs(title = paste("WY", water.yr, "relative to other years", sep = " ")) +
    my.theme +
    theme(axis.title.y = element_blank(), 
          axis.text.x  = element_text(angle=45, vjust=0.5))
  return(multiplot(p1, p2, cols = 2))
}
#-- Multiplot functino from Cookbook for R
# http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

#-- Figure Themes
# ggplot theme settings
my.theme <- theme_bw() + 
  theme(strip.background = element_rect(fill="white"), 
        strip.text = element_text(hjust = 0.1), 
        axis.title.x = element_blank())

#-- List of months
month.df <- data.frame(month = 1:12, 
                       waterMonth = c(4:12, 1:3),
                       mthName = factor(month.abb, 
                                        levels = c(month.abb[10:12], 
                                                   month.abb[1:9])))

#-- Set desired water year and month
water.yr = 2020
target.month <- filter(month.df, month == (month(today()) -1 )) %>%
  mutate(FirstDay = ymd(paste(water.yr, month, 1, sep = "-"))) %>%
  mutate(LastDay = ymd(paste(water.yr, month, days_in_month(FirstDay))))
```

```{r Data, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
#-- Load .RData
load("Climate.RData")

# Add years and months to wx.dat
wx.dat <- wx.dat %>%
  mutate(value = as.numeric(value),
         month = month(date), 
         year = year(date), 
         waterYr = ifelse(month >= 10, year + 1, year)) %>%
  left_join(month.df)
```
This document summarizes current temperature and precipitation anomalies relative to 30-year (1981-2010) normals for Canyonlands National Park during the `r water.yr` water year (October `r water.yr-1` through September `r water.yr`). 
The data were collected at the Visitor Centers by NPS personnel. 
The data were downloaded from the Global Historic Climate Network on `r format(as.Date(downloadDate, format = "%Y-%m-%d"), "%d %B, %Y")` using R (ver. `r r.ver`, R Core Team, 2020) and the rnoaa package (ver `r info$otherPkgs$rnoaa$Version`, Chamberlain, et al. 2019).  

```{r ISKY, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
#-- Filter data by desired district
my.dist <- "ISKY"
parkUnit <- filter(stations, district == my.dist) %>%
  mutate(label = paste(year(startDate), year(endDate), sep = " - "), 
         waterYr = water.yr, 
         unitName = "Island in the Sky District")
```

# `r parkUnit$unitName`
## Temperature
```{r ISKY.TempData, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
temps <- temps.dat.fn(wx.dat, my.dist, target.month$LastDay)
```
`r month.name[target.month$month]` `r water.yr` was the `r filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmin")$warmest`th warmest `r month.name[target.month$month]` in the `r round(parkUnit$timespan, 0)`-year record for `r parkUnit$unitName` (`r year(parkUnit$startDate)` to `r year(parkUnit$endDate)`).
The average daily low (TMIN) was `r abs(round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmin")$departure, 2))`°C `r ifelse(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmin")$departure > 0, "above", "below")` normal at `r round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmin")$mean.degC, 1)`°C. 
The  30-year average daily low for `r month.name[target.month$month]` is `r round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmin")$mean.30yr, 1)`°C. 
The average daily high (TMAX) was `r abs(round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmax")$departure, 2))`°C `r ifelse(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmax")$departure > 0, "above", "below")` normal at `r round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmax")$mean.degC, 1)`°C. 
The 30-year average daily high for `r month.name[month(max(temps$rank$mid.date))]` is `r round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmax")$mean.30yr, 1)`°C. 

```{r ISKY.TempFig, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
temp.fig(temps$depart)
```

\newpage
## Precipitation
```{r ISKY.PrcpData, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
prcp <- prcp.dat.fn(wx.dat, my.dist, target.month$LastDay)
```
Water year `r water.yr` is currently the `r filter(prcp$rank, waterYr == water.yr & mthName == target.month$mthName)$Driest`th driest on record out of the `r round(parkUnit$timespan, 0)`-year data set for `r parkUnit$unitName` (`r year(parkUnit$startDate)` to `r year(parkUnit$endDate)`). 
The current accumulated precipitation is `r round(filter(prcp$pct, waterYr == water.yr & mthName == target.month$mthName & element == "wy")$pct.avg, 1)`% of normal, totaling `r filter(prcp$pct, waterYr == water.yr & mthName == target.month$mthName & element == "wy")$Obs` mm of precipitation. 
The 30-year average for this time of year is `r round(filter(prcp$Avg30yr, mthName == target.month$mthName & element == "wy")$mean.30yr, 1)` mm of precipitation.

```{r ISKY.PRCPFig, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
prcp.fig(prcp$pct)
```

\newpage

```{r NEED, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
#-- Filter data by desired district
my.dist <- "NEED"
parkUnit <- filter(stations, district == my.dist) %>%
  mutate(label = paste(year(startDate), year(endDate), sep = " - "), 
         waterYr = water.yr, 
         unitName = "The Needles District")
```

# `r parkUnit$unitName`
## Temperature
```{r NEED.TempData, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
temps <- temps.dat.fn(wx.dat, my.dist, target.month$LastDay)
```
`r month.name[target.month$month]` `r water.yr` was the `r filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmin")$warmest`th warmest `r month.name[target.month$month]` in the `r round(parkUnit$timespan, 0)`-year record for `r parkUnit$unitName` (`r year(parkUnit$startDate)` to `r year(parkUnit$endDate)`).
The average daily low (TMIN) was `r abs(round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmin")$departure, 2))`°C `r ifelse(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmin")$departure > 0, "above", "below")` normal at `r round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmin")$mean.degC, 1)`°C. 
The  30-year average daily low for `r month.name[target.month$month]` is `r round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmin")$mean.30yr, 1)`°C. 
The average daily high (TMAX) was `r abs(round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmax")$departure, 2))`°C `r ifelse(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmax")$departure > 0, "above", "below")` normal at `r round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmax")$mean.degC, 1)`°C. 
The 30-year average daily high for `r month.name[month(max(temps$rank$mid.date))]` is `r round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmax")$mean.30yr, 1)`°C. 

```{r NEED.TempFig, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
temp.fig(temps$depart)
```

\newpage
## Precipitation
```{r NEED.PrcpData, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
prcp <- prcp.dat.fn(wx.dat, my.dist, target.month$LastDay)
```
Water year `r water.yr` is currently the `r filter(prcp$rank, waterYr == water.yr & mthName == target.month$mthName)$Driest`th driest on record out of the `r round(parkUnit$timespan, 0)`-year data set for `r parkUnit$unitName` (`r year(parkUnit$startDate)` to `r year(parkUnit$endDate)`). 
The current accumulated precipitation is `r round(filter(prcp$pct, waterYr == water.yr & mthName == target.month$mthName & element == "wy")$pct.avg, 1)`% of normal, totaling `r filter(prcp$pct, waterYr == water.yr & mthName == target.month$mthName & element == "wy")$Obs` mm of precipitation. 
The 30-year average for this time of year is `r round(filter(prcp$Avg30yr, mthName == target.month$mthName & element == "wy")$mean.30yr, 1)` mm of precipitation.

```{r NEED.PRCPFig, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
prcp.fig(prcp$pct)
```

\newpage

```{r MAZE, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
#-- Filter data by desired district
my.dist <- "MAZE"
parkUnit <- filter(stations, district == my.dist) %>%
  mutate(label = paste(year(startDate), year(endDate), sep = " - "), 
         waterYr = water.yr, 
         unitName = "Hans Flat Ranger Station")
```
# `r parkUnit$unitName`
## Temperature
```{r MAZE.TempData, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
temps <- temps.dat.fn(wx.dat, my.dist, target.month$LastDay)
```
`r month.name[target.month$month]` `r water.yr` was the `r filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmin")$warmest`th warmest `r month.name[target.month$month]` in the `r round(parkUnit$timespan, 0)`-year record for `r parkUnit$unitName` (`r year(parkUnit$startDate)` to `r year(parkUnit$endDate)`).
The average daily low (TMIN) was `r abs(round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmin")$departure, 2))`°C `r ifelse(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmin")$departure > 0, "above", "below")` normal at `r round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmin")$mean.degC, 1)`°C. 
The  30-year average daily low for `r month.name[target.month$month]` is `r round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmin")$mean.30yr, 1)`°C. 
The average daily high (TMAX) was `r abs(round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmax")$departure, 2))`°C `r ifelse(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmax")$departure > 0, "above", "below")` normal at `r round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmax")$mean.degC, 1)`°C. 
The 30-year average daily high for `r month.name[month(max(temps$rank$mid.date))]` is `r round(filter(temps$rank, waterYr == water.yr & mthName == target.month$mthName & element == "tmax")$mean.30yr, 1)`°C. 

```{r MAZE.TempFig, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
temp.fig(temps$depart)
```

\newpage
## Precipitation
```{r MAZE.PrcpData, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
prcp <- prcp.dat.fn(wx.dat, my.dist, target.month$LastDay)
```
Water year `r water.yr` is currently the `r filter(prcp$rank, waterYr == water.yr & mthName == target.month$mthName)$Driest`th driest on record out of the `r round(parkUnit$timespan, 0)`-year data set for `r parkUnit$unitName` (`r year(parkUnit$startDate)` to `r year(parkUnit$endDate)`). 
The current accumulated precipitation is `r round(filter(prcp$pct, waterYr == water.yr & mthName == target.month$mthName & element == "wy")$pct.avg, 1)`% of normal, totaling `r filter(prcp$pct, waterYr == water.yr & mthName == target.month$mthName & element == "wy")$Obs` mm of precipitation. 
The 30-year average for this time of year is `r round(filter(prcp$Avg30yr, mthName == target.month$mthName & element == "wy")$mean.30yr, 1)` mm of precipitation.

```{r MAZE.PRCPFig, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
prcp.fig(prcp$pct)
```
