---
title: |
  | `r park.name`
  | `r paste("Water Year", water.yr, "Climate Summary", sep = " ")`
author: "Matthew Van Scoyoc"
date: "`r Sys.Date()`"
mainfont: Arial
output:
  pdf_document: 
    fig_caption: yes
    fig_width: 6.5
    latex_engine: xelatex
editor_options:
  chunk_output_type: console
urlcolor: blue
---

```{r Setup, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
#-- Packages
# install.packages(c("tidyverse", "lubridate", "cowplot", "shadowtext", "knitr"))
library("tidyverse") # Data manipulation
library("lubridate") # Easily work with date/time data
library("cowplot") # Streamlining for ggplot2
library("shadowtext")
library("knitr") # Makes nice tables in output

#-- Custom functions
#- Number contraction
# Prints the appropraite number contraction for a given number.
number.contraction <- function(x){
  if(x == 1) {
    (y = "")
  } else if (x == 2) {
    (y = "nd")
  } else if (x == 3) {
    (y = "rd")
  } else {
    (y = "th")}
  return(y)
}

#- Number contraction
# Prints the appropraite number contraction for a given number.
visitor.center <- function(my.park){
  if(my.park == "CANY") {
    (y = "Co-op weather stations at the ISKY, MAZE, and NEED visitor centers by NPS personnel")
  } else if (my.park == "Moab") {
    (y = "Co-op weather station in downtown Moab")
  } else {
    (y = "Co-op weather station at the visitor center by NPS personnel")}
  return(y)
}

#-- Figure Theme
# ggplot theme settings
my.theme <- theme_bw() + 
  theme(strip.background = element_rect(fill="white"), 
        strip.text = element_text(hjust = 0.1), 
        axis.title.x = element_blank())

#-- Set desired park unit, water year, and month
# These are parameters hashed our as they are set in renderWaterYearReports.Rmd.
# my.park <- "MOAB"
# water.yr = 2020
```

```{r Data, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
#-- Load .RData
load("climateData.RData")

#-- Filter data by desired Park
parkUnit <- filter(stations, parkUnit == my.park) %>%
  mutate(label = paste(year(startDate), year(endDate), sep = " - "), 
         waterYr = water.yr)
```
This document summarizes current temperature and precipitation anomalies relative to 30-year (1981-2010) averages for `r parkUnit$unitName` (`r my.park`) during the `r water.yr` water year (October `r water.yr-1` through September `r water.yr`). 
The data for these analyses were collected at the `r visitor.center(my.park)` and  downloaded from the [Global Historical Climatology Network](https://www.ncdc.noaa.gov/data-access/land-based-station-data/land-based-datasets/global-historical-climatology-network-ghcn) on `r format(as.Date(downloadDate, format = "%Y-%m-%d"), "%d %B, %Y")` using R (ver. `r r.ver`, R Core Team, 2020) and the rnoaa package (ver `r info$otherPkgs$rnoaa$Version`, Chamberlain, et al. 2020). 
The data used for these analyses include the daily high temperature (TMAX), daily low temperature (TMIN), and daily precipitation accumulation (PRCP). Averages were calculated using the mean for the period of interest.
  
# Temperature
```{r TempData, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
#-- Subset temp data
temp.dat <- wx.dat %>%
  filter(element != "prcp" & parkUnit == my.park & waterYr <= water.yr) %>%
  mutate(degC = value / 10) %>%  # Convert to Celsius
  select(-value)

# -- Complete water years
# This dataframe can be used to filter out incomplete water years. 
complete.wys <- temp.dat %>%
  group_by(waterYr, mthName) %>%
  summarise(mthObs = sum(degC, na.rm = T)) %>%
  group_by(waterYr) %>%
  summarise(n = n()) %>%
  filter(n == 12)

#-- Annual average temperatures
# 30-yr average annual temperature
an.temp.30yr <- temp.dat %>%
  filter(waterYr >= 1981 & waterYr <= 2010) %>%
  group_by(parkUnit) %>%
  summarise(mean.30yr = mean(degC, na.rm = T)) %>%
  select(parkUnit, mean.30yr)
# Calculate and rank departure
an.temp.depart <- temp.dat %>%
  filter(waterYr %in% complete.wys$waterYr) %>%
  group_by(parkUnit, waterYr) %>%
  summarise(mean.degC = mean(degC, na.rm = T)) %>%
  left_join(an.temp.30yr) %>%
  mutate(departure = mean.degC - mean.30yr, 
         n = n()) %>%
  mutate(warmest = min_rank(-mean.degC),
         coolest = min_rank(mean.degC)) %>%
  arrange(waterYr)

#-- Annual annual temperatures by element (TMAX/TMIN)
# 30-yr average annual TMIN/TMAX
an.ele.30yr <- temp.dat %>%
  filter(waterYr >= 1981 & waterYr <= 2010) %>%
  group_by(element) %>%
  summarise(mean.30yr = mean(degC, na.rm = T))
# Calculate and rank departure
an.ele.depart <- temp.dat %>%
  filter(waterYr %in% complete.wys$waterYr) %>%
  group_by(element, waterYr) %>%
  summarise(mean.degC = mean(degC, na.rm = T)) %>%
  left_join(an.ele.30yr) %>%
  mutate(departure = mean.degC - mean.30yr, 
         n = n()) %>%
  arrange(element, waterYr)

#-- Monthly average temperatures 
# 30-yr Temp Stats (1981-2010 water years)
temps.30yr <- temp.dat %>%
  filter(waterYr >= 1981 & waterYr <= 2010) %>%
  group_by(element, waterMonth) %>%
  summarise(mean.30yr = mean(degC, na.rm = T)) %>%
  left_join(month.df) %>%
  select(element, mthName, mean.30yr)
# Monthly Temp Departures
temps.depart <- temp.dat %>%
  filter(waterYr %in% complete.wys$waterYr) %>%
  group_by(element, waterYr, mthName) %>%
  summarise(mean.degC = mean(degC, na.rm = T), 
                     n = n()) %>%
  left_join(temps.30yr) %>%
  mutate(departure = mean.degC - mean.30yr, 
         year = ifelse(mthName %in% c("Oct", "Nov", "Dec"), waterYr - 1, waterYr), 
         mid.date = ymd(paste(year, mthName, 15, sep = "-"))) %>%
  arrange(element, mid.date)

#-- References for text and figures
data.label <- parkUnit %>%
  mutate(label= paste(min(an.ele.depart$waterYr), 
                      max(an.ele.depart$waterYr) - 1, sep = "-"))
tmax.exceed <- temps.depart %>%
  filter(waterYr == water.yr, element == "tmax" & departure > 0)
tmin.exceed <- temps.depart %>%
  filter(waterYr == water.yr, element == "tmin" & departure > 0)
an.lm <- summary(lm(mean.degC ~ waterYr, data = an.temp.depart))$coefficients[2, 1]
tmax.lm <- summary(lm(mean.degC ~ waterYr, 
              data = filter(an.ele.depart, element == "tmax")))$coefficients[2, 1]
tmin.lm <- summary(lm(mean.degC ~ waterYr, 
              data = filter(an.ele.depart, element == "tmin")))$coefficients[2, 1]
```

Water year `r water.yr` was the `r filter(an.temp.depart, waterYr == water.yr)$warmest``r number.contraction(filter(an.temp.depart, waterYr == water.yr)$warmest)` warmest water year in the `r round(parkUnit$timespan, 0)`-year record for `r my.park` (`r year(parkUnit$startDate)` to `r year(parkUnit$endDate)`); 
the average annual temperature was `r abs(round(filter(an.temp.depart, waterYr == water.yr)$departure, 2))`°C `r ifelse(filter(an.temp.depart, waterYr == water.yr)$departure > 0, "above", "below")` the 30-year average. 
This summary suggests that temperatures are `r ifelse(an.lm > 0, "increasing", "decreasing")` `r abs(round(an.lm*10, 2))`°C per decade (figure 1).

The average annual TMAX was `r abs(round(filter(an.ele.depart, waterYr == water.yr & element == "tmax")$departure, 2))`°C `r ifelse(filter(an.ele.depart, waterYr == water.yr & element == "tmax")$departure > 0, "above", "below")` the 30-year average and has been `r ifelse(tmax.lm > 0, "increasing", "decreasing")` `r abs(round(tmax.lm*10, 2))`°C per decade (figure 2A). 
The average monthly TMAX exceeded the 30-year monthly average `r nrow(tmax.exceed)` times (`r paste(paste(tmax.exceed$mthName[1:nrow(tmax.exceed)-1], collapse = ", "), tmax.exceed$mthName[nrow(tmax.exceed)], sep = ", and ")`; figure 3A). 

The average annual TMIN was `r abs(round(filter(an.ele.depart, waterYr == water.yr & element == "tmin")$departure, 2))`°C `r ifelse(filter(an.ele.depart, waterYr == water.yr & element == "tmin")$departure > 0, "above", "below")` average and and has been `r ifelse(tmin.lm > 0, "increasing", "decreasing")` `r abs(round(tmin.lm*10, 2))`°C per decade (figure 2B). 
The monthly average TMIN exceeded the 30-year monthly average `r nrow(tmin.exceed)` times (`r paste(paste(tmin.exceed$mthName[1:nrow(tmin.exceed)-1], collapse = ", "), tmin.exceed$mthName[nrow(tmin.exceed)], sep = ", and ")`; figure 3B).

# Precipitation
```{r PrcpData, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
#-- Subset PRCP data
prcp.dat <- wx.dat %>%
  filter(element == "prcp" & parkUnit == my.park & waterYr <= water.yr) %>%
  mutate(PRCPmm = value / 10) %>% # Convert to inches
  select(-value)

# -- Complete water years
# This dataframe can be used to filter out incomplete water years. 
complete.wys <- prcp.dat %>%
  group_by(waterYr, mthName) %>%
  summarise(mthObs = sum(PRCPmm, na.rm = T)) %>%
  group_by(waterYr) %>%
  summarise(n = n()) %>%
  filter(n == 12)

#-- Monthly departure
# Calculate 30-year monthly averages
mth.30yr <- prcp.dat %>%
  filter(waterYr >= 1981 & waterYr <= 2010) %>%
  group_by(waterYr, mthName) %>%
  summarise(Obs = sum(PRCPmm, na.rm = T)) %>%
  group_by(mthName) %>%
  summarise(mean.30yr = mean(Obs, na.rm = T))
# Calculate monthly precip totals and departure
mth.prcp <- prcp.dat %>%
  filter(waterYr %in% complete.wys$waterYr) %>%
  group_by(waterYr, mthName) %>%
  summarise(Obs = sum(PRCPmm, na.rm = T)) %>%
  left_join(mth.30yr) %>%
  mutate(pct.avg = (Obs / mean.30yr) * 100,
         departure = Obs - mean.30yr) %>%
  group_by(mthName) %>%
  mutate(driest = min_rank(Obs))

#-- Water Year departure (by month)
# Calculate 30-year monthly averages
wy.30yr <- prcp.dat %>%
  filter(waterYr >= 1981 & waterYr <= 2010) %>%
  group_by(waterYr, mthName) %>%
  summarise(Obs = sum(PRCPmm, na.rm = T)) %>%
  mutate(Obs = cumsum(Obs)) %>%
  group_by(mthName) %>%
  summarise(mean.30yr = mean(Obs))
# Calculate monthly precip totals and departure
wy.prcp <- prcp.dat %>%
  filter(waterYr %in% complete.wys$waterYr) %>%
  group_by(waterYr, mthName) %>%
  summarise(Obs = sum(PRCPmm, na.rm = T)) %>%
  mutate(Obs = cumsum(Obs)) %>%
  left_join(wy.30yr) %>%
  mutate(pct.avg = (Obs / mean.30yr) * 100,
         departure = Obs - mean.30yr) %>%
  group_by(mthName) %>%
  mutate(driest = min_rank(Obs))

#-- Water Year departure (total for water year)
# Calculate 30-year monthly averages
wy.30yr <- prcp.dat %>%
  filter(waterYr >= 1981 & waterYr <= 2010) %>%
  group_by(waterYr) %>%
  summarise(Obs = sum(PRCPmm, na.rm = T)) %>%
  ungroup() %>%
  summarise(mean.30yr = mean(Obs))
# Filter water year totals and assign rank
wy.prcp.tot <- prcp.dat %>%
  filter(waterYr %in% complete.wys$waterYr) %>%
  group_by(waterYr) %>%
  summarise(Obs = sum(PRCPmm, na.rm = T)) %>%
  mutate(mean.30yr = wy.30yr$mean.30yr,
         departure = Obs - mean.30yr,
         pct.avg = (Obs / mean.30yr) * 100,
         driest = min_rank(Obs))

#-- References for text and figures
mth.above <- filter(mth.prcp, waterYr == water.yr & departure > 0)
mth.below <- filter(mth.prcp, waterYr == water.yr & departure < 0)
prcp.lm <- summary(lm(Obs ~ waterYr, data = wy.prcp.tot))$coefficients[2, 1]
```

Water year `r water.yr` was the `r filter(wy.prcp.tot, waterYr == water.yr)$driest``r number.contraction(filter(wy.prcp.tot, waterYr == water.yr)$driest)` driest on record for `r my.park`. 
Total precipitation was `r abs(round(filter(wy.prcp.tot, waterYr == water.yr)$departure, 1))` mm less than the 30-year average, measuring `r filter(wy.prcp.tot, waterYr == water.yr)$Obs` mm of precipitation. 
This summary suggests that precipitation is `r ifelse(prcp.lm > 0, "increasing", "decreasing")` at a rate of `r abs(round(prcp.lm*10, 2))` mm per decade (figures 4A).

`r paste(paste(mth.above$mthName[1:nrow(mth.above)-1], collapse = ", "), mth.above$mthName[nrow(mth.above)], sep = ", and ")` received above average precipitation (figures 4B), but were not enough to compensate for the `r nrow(mth.below)` months that were below average (`r paste(paste(mth.below$mthName[1:nrow(mth.below)-1], collapse = ", "), mth.below$mthName[nrow(mth.below)], sep = ", and ")`; figure 4C).

\newpage
# Figures
```{r Fig1, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.cap="Average annual temperatures."}
#-- Average WY Temps
# Labels
t1.labs <- an.temp.30yr %>%
  mutate(label = paste0("1981-2010 Mean: ", round(mean.30yr, 1), "°C"))
# Figure
t1 <- ggplot(an.temp.depart, aes(x = waterYr, y = mean.degC)) +
  geom_ribbon(aes(ymin = min(mean.degC) - 0.1, ymax = mean.degC), 
              stat = "identity", fill = "orange", alpha = 0.5) +
  geom_hline(aes(yintercept = an.temp.30yr$mean.30yr), linetype="dashed", 
             size = 0.5, color = "black") +
  geom_line(stat = "identity", color = "red3", size = 1) +
  geom_point(stat = "identity", color = "red3", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_shadowtext(data = t1.labs, color = "black", bg.colour = "white", 
                  mapping = aes(x = -Inf, y = mean.30yr, label = label), 
                  size = 3, hjust = -0.1, vjust = -0.5) +
  labs(y = "Temperature (°C)") +
  my.theme
t1
```

```{r Fig2, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.cap="Average annual TMAX and TMIN."}
#-- Average WY TMAX/TMIN
# Labels
t2.labs <- an.ele.30yr %>%
  mutate(label = paste0("1981-2010 Mean: ", round(mean.30yr, 1), "°C"))
# Add y-axis limits to data frame for plotting
an.ele.depart <- an.ele.depart %>%
  left_join(ungroup(an.ele.depart) %>%
              group_by(element) %>%
              summarise(ylim = min(mean.degC)))
# Figure
t2 <- ggplot(an.ele.depart, aes(x = waterYr, y = mean.degC)) +
  geom_ribbon(aes(ymin = ylim - 0.2, ymax = mean.degC), 
              stat = "identity", fill = "orange", alpha = 0.5) +
  geom_hline(data = an.ele.30yr, mapping = aes(yintercept = mean.30yr), 
             linetype="dashed", size = 0.5, color = "black") +
  geom_line(stat = "identity", color = "red3", size = 1) +
  geom_point(stat = "identity", color = "red3", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_shadowtext(data = t2.labs, color = "black", bg.colour = "white", 
                  mapping = aes(x = -Inf, y = mean.30yr, label = label), 
                  size = 3, hjust = -0.1, vjust = -0.5) +
  facet_wrap(~ element, nrow = 2,  scales = "free_y",
             labeller = as_labeller(c(tmax = "A. Annual average TMAX", 
                                      tmin = "B. Annual average TMIN"))) +
  labs(y = "Temperature (°C)") +
  my.theme +
  theme(axis.text.x  = element_text(angle=45, vjust=0.5))
t2
```

```{r Fig3, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.cap="Average monthly TMAX and TMIN."}
#-- Monthly departure TMAX/TMIN
t3 <- ggplot(temps.depart, aes(x = mthName, y = departure, group = waterYr)) +
  geom_line(stat = "identity", color = "gray") +
  geom_shadowtext(data = data.label, color = "gray40", bg.colour = "white", 
                  mapping = aes(x = Inf, y = -Inf, label = label), 
                  hjust = 1.4, vjust = -1) +
  geom_hline(aes(yintercept = 0), linetype="dashed", size = 0.5, 
             color = "black") +
  geom_line(data = filter(temps.depart, waterYr == water.yr), stat = "identity", 
            color = "red3", size = 1) +
  geom_point(data = filter(temps.depart, waterYr == water.yr), stat = "identity", 
            color = "red3", size = 2) +
  geom_shadowtext(data = data.label, color = "red3", bg.colour = "white", 
                  mapping = aes(x = Inf, y = Inf, label = waterYr), 
                  hjust = 2, vjust = 1.5, fontface="bold") +
  facet_wrap(~ element, nrow = 2, scales = "free_y",
             labeller = as_labeller(c(tmax = "A. Montly average TMAX", 
                                      tmin = "B. Montly average TMIN"))) +
  labs(y = "Temperature (°C)") +
  my.theme +
  theme(axis.text.x  = element_text(angle=45, vjust=0.5))
t3
# #-- Combine figure into one big figure
# bottom.row <- plot_grid(t2, t3, labels = c("A.", "B."), ncol = 2)
# plot_grid(t1, bottom.row, labels = c("A.", ""), ncol = 1, scale = c(0.8, 1))
```

```{r Fig4, echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.cap="Water year precipitation totals and monly departures from 30-year averages."}
#-- Water year precipitation totals
# Labels
wy.30yr <- wy.30yr %>%
  mutate(label = paste0("1981-2010 Mean: ", round(mean.30yr, 1), "mm"))
# Figure
p1 <- ggplot(wy.prcp.tot, aes(x = waterYr, y = Obs)) +
  geom_area(stat = "identity", fill = "forestgreen", alpha = 0.5) +  
  geom_hline(aes(yintercept = wy.30yr$mean.30yr), linetype="dashed", 
             size = 0.5, color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_line(stat = "identity", color = "darkgreen", size = 1) +
  geom_point(stat = "identity", color = "darkgreen", size = 2) +
  geom_shadowtext(data = wy.30yr, color = "black", bg.colour = "white", 
                  mapping = aes(x = -Inf, y = mean.30yr, label = label), 
                  size = 3, hjust = -0.1, vjust = 1.5) +
  labs(y = "Precipitation (mm)", title = "Water year precipitation totals") +
  my.theme

# Monthly precipitation totals
p2 <- ggplot(mth.prcp, aes(x = mthName, y = departure, group = waterYr)) +
  geom_line(stat = "identity", color = "gray") +
  geom_shadowtext(data = data.label, color = "gray40", bg.colour = "white", 
                  mapping = aes(x = -Inf, y = Inf, label = label),
                  hjust = -0.5, vjust = 1.5) +
  geom_hline(aes(yintercept = 0), linetype="dashed", size = 0.5, 
             color = "black") +
  geom_line(data = filter(mth.prcp, waterYr == water.yr), 
            stat = "identity", color = "darkgreen", size = 1) +
  geom_point(data = filter(mth.prcp, waterYr == water.yr), 
            stat = "identity", color = "darkgreen", size = 2) +
  geom_shadowtext(data = data.label, color = "darkgreen", bg.colour = "white", 
                  mapping = aes(x = Inf, y = Inf, label = waterYr), 
                  hjust = 2.5, vjust = 1.5, fontface="bold") +
  labs(y = "Precipitation (mm)", 
       title = "Monthly departure relative to other water years") +
  my.theme +
  theme(axis.text.x  = element_text(vjust=0.5))

# Water year accumulation
p3 <- ggplot(wy.prcp, aes(x = mthName, y = departure, group = waterYr)) +
  geom_line(stat = "identity", color = "gray") +
  geom_shadowtext(data = data.label, color = "gray40", bg.colour = "white", 
                  mapping = aes(x = -Inf, y = Inf, label = label),
                  hjust = -0.5, vjust = 1.5) +
  geom_hline(aes(yintercept = 0), linetype="dashed", size = 0.5, 
             color = "black") +
  geom_line(data = filter(wy.prcp, waterYr == water.yr), 
            stat = "identity", color = "darkgreen", size = 1) +
  geom_point(data = filter(wy.prcp, waterYr == water.yr), 
            stat = "identity", color = "darkgreen", size = 2) +
  geom_shadowtext(data = data.label, color = "darkgreen", bg.colour = "white", 
                  mapping = aes(x = Inf, y = Inf, label = waterYr), 
                  hjust = 2.5, vjust = 1.5, fontface="bold") +
  labs(y = "Precipitation (mm)", 
       title = "Accumulating departure through the water year") +
  my.theme +
  theme(axis.text.x  = element_text(vjust=0.5))

# Using cowplot to arrange figures
plot_grid(p1, p2, p3, labels = c("A.", "B.", "C."), ncol = 1)
```
